---
title: "Using acsmoe with tidycensus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using acsmoe with tidycensus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`tidycensus` is the right tool for downloading ACS estimates and margins of
error. `acsmoe` starts after that: it works with estimate/MOE columns that you
already have.

Kyle Walker's `tidycensus` article on ACS margins of error shows the standard
workflow and the standard Census approximation formulas:
<https://walker-data.com/tidycensus/articles/margins-of-error.html>.

That article demonstrates `tidycensus::moe_sum()`, `tidycensus::moe_prop()`,
`tidycensus::moe_ratio()`, and `tidycensus::moe_product()`. It also quotes the
Census warning that these approximation methods do not account for correlation
or covariance between basic estimates. `acsmoe` is intended for the same
tabular ACS regime, but exposes covariance-aware extensions and a grouped
aggregation helper.

## Pull data with tidycensus

This example mirrors the shape of Walker's MOE vignette: pull tract-level ACS
age-by-sex cells, then aggregate those cells into a derived total. The call is
not evaluated in this vignette because it requires network access and, in many
setups, a Census API key.

```{r, eval = FALSE}
library(tidycensus)
library(dplyr)
library(acsmoe)

vars <- paste0("B01001_0", c(20:25, 44:49))

ramsey <- get_acs(
  geography = "tract",
  variables = vars,
  state = "MN",
  county = "Ramsey",
  year = 2016
)

ramsey65 <- ramsey |>
  group_by(GEOID) |>
  summarize(
    estimate_65plus = sum(estimate),
    moe_65plus = acs_sum(estimate, moe)$moe,
    .groups = "drop"
  )
```

With no covariance supplied, `acs_sum()` intentionally reduces to the same
zero-covariance root-sum-square calculation used by `tidycensus::moe_sum()`.
That makes it a drop-in bridge from the standard workflow to more explicit
uncertainty propagation.

The package website includes a fuller `tidycensus` example with evaluated maps
when the site is built with Census API credentials. That example is kept out of
CRAN vignette evaluation because it requires network access, `sf` geometries,
and current ACS API availability.

## Work from paired estimate/MOE columns

Many ACS workflows become wide after `tidycensus::get_acs(output = "wide")`, or
after a user-created join. `acs_aggregate()` handles this paired-column form.

```{r}
library(acsmoe)

tracts <- data.frame(
  region = c("north", "north", "south", "south"),
  population = c(1000, 1200, 900, 1100),
  population_moe = c(120, 140, 100, 130),
  households = c(420, 500, 360, 440),
  households_moe = c(60, 70, 50, 65)
)

acs_aggregate(
  tracts,
  group_var = "region",
  value_cols = c("population", "households"),
  moe_cols = c("population_moe", "households_moe")
)
```

The default `cov_strategy = "zero"` is deliberately conservative in the API
sense: it matches the standard Census approximation behavior. It should not be
read as a claim that tract estimates are truly independent.

## Add covariance when you have it

If a covariance matrix is available from an external method, pass it on the
standard-error scale. Do not pass covariance of MOEs.

```{r}
estimates <- c(1000, 1200)
moes <- c(120, 140)
ses <- moe_to_se(moes)

cov_mat <- matrix(
  c(ses[1]^2, 1500,
    1500, ses[2]^2),
  nrow = 2
)

acs_sum(estimates, moes, cov = cov_mat)
```

For aggregation, `cov_strategy = "constant"` accepts a scalar correlation and
constructs a valid covariance matrix from the input MOEs. This is useful for
sensitivity analysis, not as an automatic estimator of ACS covariance.

```{r}
acs_aggregate(
  tracts,
  group_var = "region",
  value_cols = "population",
  moe_cols = "population_moe",
  cov_strategy = "constant",
  cov_value = 0.25
)
```

## What this package does not do

`acsmoe` does not download ACS data. Use `tidycensus` for that.

`acsmoe` does not estimate variance from microdata. Use `survey` or `srvyr` for
PUMS and replicate-weight workflows.

`acsmoe` also does not implement regionalization. Walker's `tidycensus` MOE
article points readers to Spielman and Folch's regionalization work and an old
Python implementation. That historical code lives at
<https://github.com/geoss/censumander>. We used it as development-only
reference material for formula checks, but regionalization itself is out of
scope for this package.

The boundary is intentional: `acsmoe` focuses on propagation of uncertainty for
tabular estimate/MOE workflows after ACS data have already been obtained.

## References

- U.S. Census Bureau. 2020. *Understanding and Using American Community Survey
  Data: What All Data Users Need to Know*. See Chapter 8, "Calculating Measures
  of Error for Derived Estimates."
  <https://www.census.gov/programs-surveys/acs/library/handbooks/general.html>
- Walker, Kyle, and Matt Herman. 2025. `tidycensus`: Load US Census Boundary
  and Attribute Data as `tidyverse` and `sf`-Ready Data Frames.
  <https://CRAN.R-project.org/package=tidycensus>
- Walker, Kyle. "Margins of error in the ACS."
  <https://walker-data.com/tidycensus/articles/margins-of-error.html>
- Spielman, Seth E., David Folch, and Nicholas Nagle. 2014. "Patterns and
  Causes of Uncertainty in the American Community Survey." *Applied Geography*
  46: 147-157. <https://doi.org/10.1016/j.apgeog.2013.11.002>
- Spielman, Seth E., and David C. Folch. 2015. "Reducing Uncertainty in the
  American Community Survey through Data-Driven Regionalization." *PLOS ONE*
  10(2): e0115626. <https://doi.org/10.1371/journal.pone.0115626>
- Folch, David C., Daniel Arribas-Bel, Julia Koschinsky, and Seth E. Spielman.
  2016. "Spatial Variation in the Quality of American Community Survey
  Estimates." *Demography* 53(5): 1535-1554.
  <https://doi.org/10.1007/s13524-016-0499-1>
- Folch, David C., and Seth E. Spielman. `geoss/censumander`. Historical Python
  reference implementation used here only for development validation.
  <https://github.com/geoss/censumander>
