---
title: "Mapping ACS uncertainty with tidycensus"
description: >
  A pkgdown-only worked example using live tidycensus data, ACS MOEs, and maps.
---

```{r, include = FALSE}
live_examples <- identical(Sys.getenv("ACSMOE_LIVE_EXAMPLES"), "true") &&
  nzchar(Sys.getenv("CENSUS_API_KEY")) &&
  requireNamespace("tidycensus", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = live_examples,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

This article is intended for the pkgdown site, not CRAN checks. It evaluates
only when the site build sets `ACSMOE_LIVE_EXAMPLES=true` and provides a
`CENSUS_API_KEY`.

```{r, eval = !live_examples, echo = FALSE}
cat(
  "Live examples are disabled in this build. ",
  "Set ACSMOE_LIVE_EXAMPLES=true and CENSUS_API_KEY to render the maps."
)
```

The example uses `tidycensus` to download tract geometries and ACS estimate/MOE
pairs, then uses `acsmoe` to propagate uncertainty and map the result. The
workflow follows the margin-of-error framing in Kyle Walker's `tidycensus`
article: <https://walker-data.com/tidycensus/articles/margins-of-error.html>.

## Get tract data

```{r setup}
library(acsmoe)
library(tidycensus)
library(dplyr)
library(sf)
library(ggplot2)

tidycensus::census_api_key(Sys.getenv("CENSUS_API_KEY"), install = FALSE)
```

We will estimate the population age 65 and older by summing age-by-sex cells
from table `B01001`. The ACS API returns a row per variable per tract, with an
estimate and a margin of error for each row.

```{r get-data}
vars_65plus <- paste0("B01001_", sprintf("%03d", c(20:25, 44:49)))

ramsey_age <- get_acs(
  geography = "tract",
  variables = vars_65plus,
  state = "MN",
  county = "Ramsey",
  year = 2022,
  survey = "acs5",
  geometry = TRUE
)
```

## Propagate tract-level uncertainty

With no covariance supplied, `acs_sum()` matches the standard Census
root-sum-square MOE approximation used by `tidycensus::moe_sum()`. The benefit
of using `acsmoe` here is that the same code path can later accept covariance
when you have it.

```{r tract-propagation}
ramsey_65plus <- ramsey_age |>
  group_by(GEOID, NAME) |>
  summarize(
    estimate_65plus = sum(estimate),
    moe_65plus = acs_sum(estimate, moe)$moe,
    .groups = "drop"
  ) |>
  mutate(
    cv_65plus = acs_cv(estimate_65plus, moe_65plus),
    relative_moe = moe_65plus / estimate_65plus,
    reliability = acs_reliability(estimate_65plus, moe_65plus)
  )

ramsey_65plus |>
  st_drop_geometry() |>
  select(GEOID, estimate_65plus, moe_65plus, cv_65plus, reliability) |>
  head()
```

## Map estimates and uncertainty

The estimate map is useful, but it is incomplete on its own. The relative-MOE
and reliability maps show where the tract estimates are more or less stable.

```{r estimate-map}
ggplot(ramsey_65plus) +
  geom_sf(aes(fill = estimate_65plus), color = "white", linewidth = 0.1) +
  scale_fill_gradient(
    low = "#f4f1de",
    high = "#245953",
    labels = scales::comma,
    name = "Age 65+"
  ) +
  labs(
    title = "Estimated population age 65+",
    subtitle = "Ramsey County, MN tracts; ACS 2018-2022 5-year"
  ) +
  theme_void()
```

```{r relative-moe-map}
ggplot(ramsey_65plus) +
  geom_sf(aes(fill = relative_moe), color = "white", linewidth = 0.1) +
  scale_fill_gradient(
    low = "#e8f3f1",
    high = "#a23e48",
    labels = scales::percent,
    name = "MOE / estimate"
  ) +
  labs(
    title = "Relative uncertainty in age 65+ estimates",
    subtitle = "Higher values indicate less stable tract estimates"
  ) +
  theme_void()
```

```{r reliability-map}
ggplot(ramsey_65plus) +
  geom_sf(aes(fill = reliability), color = "white", linewidth = 0.1) +
  scale_fill_manual(
    values = c(
      reliable = "#2a9d8f",
      caveat = "#e9c46a",
      unreliable = "#e76f51"
    ),
    drop = FALSE,
    name = "CV class"
  ) +
  labs(
    title = "Reliability classes from coefficient of variation",
    subtitle = "Default thresholds: reliable < 0.12, caveat < 0.40"
  ) +
  theme_void()
```

## Aggregate to custom regions

Users often aggregate tracts into service areas, planning districts, or
hand-built regions. Here we create a simple west/east split from tract
centroids only to demonstrate the mechanics.

```{r aggregate-regions}
centroids <- st_coordinates(st_centroid(ramsey_65plus))

ramsey_regions <- ramsey_65plus |>
  mutate(region = if_else(centroids[, "X"] < median(centroids[, "X"]), "west", "east"))

region_uncertainty <- acs_aggregate(
  st_drop_geometry(ramsey_regions),
  group_var = "region",
  value_cols = "estimate_65plus",
  moe_cols = "moe_65plus"
) |>
  mutate(
    cv_65plus = acs_cv(estimate_65plus, moe_65plus),
    relative_moe = moe_65plus / estimate_65plus
  )

region_shapes <- ramsey_regions |>
  group_by(region) |>
  summarize(.groups = "drop") |>
  left_join(region_uncertainty, by = "region")

region_uncertainty
```

```{r aggregate-map}
ggplot(region_shapes) +
  geom_sf(aes(fill = relative_moe), color = "white", linewidth = 0.4) +
  geom_sf_text(aes(label = region), color = "white", fontface = "bold") +
  scale_fill_gradient(
    low = "#31572c",
    high = "#bc4749",
    labels = scales::percent,
    name = "MOE / estimate"
  ) +
  labs(
    title = "Aggregated uncertainty for custom regions",
    subtitle = "Toy west/east split, shown only as an aggregation example"
  ) +
  theme_void()
```

## Sensitivity to covariance assumptions

The default aggregation assumes zero covariance, matching the standard Census
approximation. If you want a sensitivity analysis, `cov_strategy = "constant"`
lets you compare against a simple common-correlation assumption. This does not
estimate ACS covariance; it only propagates an assumption you explicitly pass.

```{r covariance-sensitivity}
zero_cov <- acs_aggregate(
  st_drop_geometry(ramsey_regions),
  group_var = "region",
  value_cols = "estimate_65plus",
  moe_cols = "moe_65plus",
  cov_strategy = "zero"
) |>
  mutate(strategy = "zero covariance")

positive_corr <- acs_aggregate(
  st_drop_geometry(ramsey_regions),
  group_var = "region",
  value_cols = "estimate_65plus",
  moe_cols = "moe_65plus",
  cov_strategy = "constant",
  cov_value = 0.20
) |>
  mutate(strategy = "constant correlation = 0.20")

bind_rows(zero_cov, positive_corr) |>
  mutate(relative_moe = moe_65plus / estimate_65plus) |>
  select(region, strategy, estimate_65plus, moe_65plus, relative_moe)
```

The important distinction is methodological: `acsmoe` can propagate covariance
when supplied, but it does not infer a covariance structure from ACS tabular
MOEs. That inference problem is separate from the mechanics of propagation.
